DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2016-06-02 2015CompleteDerivation/completeDerivation.tex
Christine's comments on classical hamilotnian

	Consistently with what we said in Assumption \ref{detrevass}, if we defined a state for each particle, then a deterministic and reversible evolution on such state must exist. Yet, the idea that we can assign states to infinitesimal parts should be considered only a \emph{simplifying} assumption. The obvious reason is that we know this does not work in practice: as we keep decomposing the material we end up with elementary particles. But it is more instructive to understand when and how the assumption breaks down at a more conceptual level.

@@

	If we just assume the whole system is deterministic and reversible, then each part can evolve depending the states of the other parts. But this does not solve the problem entirely. Let's ignore the question of how we were able to define the states of the parts. Let's also ignore that the type of forces one would typically use (i.e. ones that conserve energy or ones with scalar and vector potentials) are precisely the ones we derive by assuming the pieces are not interacting with each other. The problem is that we can always locally separate the evolution into independent degrees of freedom. For example, the position and momentum of two particles may affect each other during the evolution, but the average and difference in position and momentum may evolve independently.\footnote{Also note that any Hamiltonian is locally isomorphic to a free particle (citation needed)} We can always find such local decomposition, and the easiest way to see that is in terms of the transported variables: they retain the original value, they clearly evolve independently. And since the pieces are infinitesimal, a local decomposition is all that is needed. While such description may be cumbersome to achieve in practice, conceptually it is still possible. So, even if the particles are in principle interacting, the system is described by degrees of freedom that evolve independently, oblivious to each other.

@@

\begin{prop}\label{canonical_transformation}
	A deterministic and reversible evolution map for a homogeneous classical material is self-symplectomorphism. That is: $\mathcal{T}_{\Delta t}: T^*\mathcal{Q} \rightarrow T^*\mathcal{Q}$ and $\mathcal{T}_{\Delta t}^*\omega = \omega$ where $\mathcal{T}_{\Delta t}^*$ is the pullback of $\mathcal{T}_{\Delta t}$.
\end{prop}

\begin{justification}
	We claim $\mathcal{T}_{\Delta t}$ is a self-homeomorphism on $T^*\mathcal{Q}$. The state space for particles of a homogeneous classical material is $T^*\mathcal{Q}$ by \ref{classical_phase_space} $\mathcal{T}_{\Delta t}$ is a deterministic and reversible evolution map and \ref{detrevmap}.

@@

In the special case of continuous time evolution we can prove the following:

\begin{prop}\label{hamiltonian}
	A continuous deterministic and reversible process for a homogeneous classical material admits a Hamiltonian $H: T^*\mathcal{Q} \rightarrow \mathbbm{R}$ that allows to write the laws of evolution as
\begin{align*}
d_{t}q^i &= \partial_{p_i} H \\
d_{t}p_i &= - \partial_{q^i} H
\end{align*}
\end{prop}

@@

We claim the state variables evolve according to Hamilton's equations. $S^{a} \omega_{ab} = d_t\xi^a \omega_{ab} = \partial_{b}H$. For $b=\{1,...,n\}$ we have $d_tp_i (-1) = \partial_{q_i} H$. For $a=\{n+1,...,2n\}$ we have $d_tq^i (+1) = \partial_{p_i} H$.
\end{justification}

We recognize the familiar set of Hamilton's equations. They are the set of equation to describe the deterministic and reversible motion of the infinitesimal parts of an infinitesimally reducible homogeneous material. In other words: the forces that conserve energy (i.e. the value of a suitable Hamiltonian) are exactly the ones that provide deterministic and reversible motion. The challenge in their derivation mainly lies in the necessary use of different branches of mathematics (e.g. topology, measure theory, differential geometry, symplectic geometry). The conceptual meaning, on the other hand, is hopefully straight forward.

It's important to realize that, during the derivation, multiple mathematical features (e.g. invariant densities, cotangent bundle for phase space, symplectomorphism) where justified by the same physical assumption. As such, the math itself does not give us a conceptually unified picture, it gets lost. This is one of the reason we believe that starting from the physical description is objectively better if we are to come to a better understanding of our physical theories.

Note that we could have taken different approaches. For example, we could have appealed to information theory and asked our invariant distributions to preserve Shannon's information entropy, as no information should be gained or lost during a deterministic and reversible process. Or we could have appealed to statistical mechanics and required that the determinant of the covariance matrix be conserved, as deterministic and reversible process should be defined at the same level of precision. With suitable treatment of independent d.o.f. both these approaches recover Hamiltonian mechanics as well. While the full treatment goes outside the scope of this work, we want to underline that the definitions used in this work go a long way to building bridges at the core of different mathematical and scientific disciplines.

@@

The discussion so far has been limited to the case where both the set of states and the evolution map are defined at equal time and never change throughout the evolution. This is too restrictive as there are very reasonable situation in which this does not hold.

The first obvious case is that is that the map may be time dependent. Note that this does not affect our definition of determinism and reversibility: we still can tell final state from initial state and vice-versa. We just need to specify the time as well. At this point we have no way to do that. For example, the Hamiltonian $H: T^*\mathcal{Q} \rightarrow \mathbbm{R}$ is just a function of the state.

@@

The third case is when we perform a state variable transformation $q'=q'(q,t)$ that is time dependent (e.g. $x'=x+vt$). Our composite state distribution $\rho(q)$ becomes $\rho(q',t)$: it is no longer defined at equal time. This means that both our measure $\mu$ and our metric $d\mu$ need to be modified.

In all three cases we can't simply stick time in the Hamiltonian and expect everything to works out. But we also can't ignore the problem either as the three cases outlined are pretty common situations to study. So we need to go back to our definitions and amend them properly.

@@

The first thing to do is to extend our $n$-dimensional $\mathcal{Q}$, the manifold that defines our units, to include the time variable $t$. We call $\mathcal{M}$ such extended space which is a manifold of dimension $n+1$: state variables and time variable are independent (i.e. any combination of state and time is valid). A time variable is not simply another state variables: they don't identify extra states and it's only on the former that we have defined deterministic and reversible evolution. When changing variables, we still need to make sure that deterministic and reversible motion can be defined on the new time variable. Transformations like $t'=x \; x'=t$ (exchanging time with another state variable) or $t'=\sqrt{t^2 + q^2} \; x'=\tan^{-1}(t/x)$ (polar coordinates between time and state variable) clearly do not guarantee deterministic and reversible motion over the new time variable. The change of time variable must preserve the ordering therefore the transformation has to be strictly monotonic. In this sense, the time variable is special.

As before, we are interested in invariant densities which are defined on co-vectors. The time variable $t$ will have a conjugate quantity $E\equiv\hbar\bar{\omega}$.\footnote{$\bar{\omega}$ is the classical analogue of the wave frequency. We use $\bar{\omega}$ to distinguish from the phase-space metric $\omega$.} We call the combination of $(t, E)$ the temporal degree of freedom. As time is special, so is the temporal degree of freedom, which is treated differently when defining distribution $\rho_\mathcal{c}$ and therefore by the measure $\mu$ and the metric $\omega$.

The first thing to note is that $E$ cannot be an independent variable. It cannot add any states, as those are defined by $T^*\mathcal{Q}$ only, and cannot add any time instants, as those are defined by $t$ alone. Therefore there must exist a constraint such that, locally, $E=E(t,q^i,p_i)$. The set of all states at all time is a sub-manifold of $T^*\mathcal{M}$ of dimension $2n+1$.

@@

We can understand the geometry of $T^*\mathcal{M}$ in the following way. As we have seen, the temporal degree of freedom $(t, E)$ and the other d.o.f. $(q^i, p_i)$ are not independent. So they are not orthogonal in the extended phase space. States are defined on the plane where $\langle q, p \rangle$ (maximally) change: this is not the plane of constant $\langle t, E \rangle$ (they are not orthogonal) where $dq \wedge dp$ is defined, but the plane perpendicular to constant $\langle q, p \rangle$ where $dt \wedge dE$ is defined. It is on that plane we can properly count states and define our metric.

@@

The metric $\omega$ is still the canonical two form expressed with the mathematically non-canonical, but physically meaningful, variable $E$. This allows us to better understand its physical meaning. The metric allows to count the possibilities of within a degree of freedom adjusting the metric on $T^*\mathcal{Q}$ to avoid the counting problems introduced by time variable changes.

The metric actually fixes two other problems. The first is: how do we distinguish temporal from standard d.o.f.? Given a two-dimensional surface $U \subset T^*\mathcal{M}$, how can we tell if it should be charted by a $(t,E)$ or whether we can define a distribution on it? Consider $\int_U \omega$. The result will be positive if the integration is over a standard degree of freedom, and it will be negative if the integration is over the temporal degree of freedom. If the contribution in any subregion of $U$ is positive, then the whole $U$ is always oriented along a standard degree of freedom.

The second problem is: how do we prevent changes in time variable that disturb determinism and reversibility? The invariance of $\omega$ already guarantees that. As we saw before the vector product is conserved within a degree of freedom is equivalent to saying that changing variables conserves the possibilities within a d.o.f. In fact, the flow of an infinitesimal transformation that preserve the vector product an a surface is a divergence free field. The conservation of the Minkowsky product across the temporal and standard d.o.f. means that changing variables does not introduce rotation between time and state variables. In face, the flow of an infinitesimal transformation that preserve the Minkowsky product on a surface is a curl free field.

We can capture the above discussion by stating that the extended state space is an hypersurface of the symplectic manifold $(T^*\mathcal{M}, \omega)$, where $\omega \equiv \sum dq^i \wedge dp_i - dt \wedge dE$. This is the simplest manifold that allows us to define time-and-state-variable-invariant densities. The metric allows us to measure the cardinality of possibilities.

\begin{prop}\label{classical_extended_phase_space}
	The extended state space, the set of states at all possible times, for the particles of a homogeneous classical material is an hyper-surface of the symplectic manifold formed by a cotangent bundle $T^*\mathcal{M}$ equipped with the canonical two-form $\omega = \sum dq^i \wedge dp_i - dt \wedge dE$.
\end{prop}

>>>>

	Consistently with what we said in Assumption \ref{detrevass}, if we defined a state for each particle, then a deterministic and reversible evolution on such state must exist. Yet, the idea that we can assign states to infinitesimal parts should be considered only a \emph{simplifying} assumption. The obvious reason is that we know this does not work in practice: as we keep decomposing the material we end up with molecules, atoms and then subatomic particles. But it is more instructive to understand when and how the assumption breaks down at a more conceptual level.

@@

	If we just assume the whole system is deterministic and reversible, then each part can evolve depending on the states of the other parts. But this does not solve the problem entirely. Let's ignore the question of how we were able to define the states of the parts. Let's also ignore that the type of forces one would typically use (i.e. ones that conserve energy or ones with scalar and vector potentials) are precisely the ones we derive by assuming the pieces are not interacting with each other. The problem is that we can always locally separate the evolution into independent degrees of freedom. For example, the position and momentum of two particles may affect each other during the evolution, but the average and difference in position and momentum may evolve independently.\footnote{Also note that any Hamiltonian is locally isomorphic to a free particle (citation needed)} We can always find such local decomposition, and the easiest way to see that is in terms of the transported variables: they retain the original value, they clearly evolve independently. And since the pieces are infinitesimal, a local decomposition is all that is needed. While such description may be cumbersome to achieve in practice, conceptually it is still possible. So, even if the particles are in principle interacting, the system is described by degrees of freedom that evolve independently, oblivious to each other.

@@

\begin{prop}\label{canonical_transformation}
	A deterministic and reversible evolution map for a homogeneous classical material is a self-symplectomorphism. That is: $\mathcal{T}_{\Delta t}: T^*\mathcal{Q} \rightarrow T^*\mathcal{Q}$ and $\mathcal{T}_{\Delta t}^*\omega = \omega$ where $\mathcal{T}_{\Delta t}^*$ is the pullback of $\mathcal{T}_{\Delta t}$.
\end{prop}

\begin{justification}
	We claim $\mathcal{T}_{\Delta t}$ is a self-homeomorphism on $T^*\mathcal{Q}$. The state space for particles of a homogeneous classical material is $T^*\mathcal{Q}$ by \ref{classical_phase_space}. $\mathcal{T}_{\Delta t}$ is a deterministic and reversible evolution map and by \ref{detrevmap} is a self-homeomorphism.

@@

If we assume continuous time evolution we have the following:

\begin{prop}\label{hamiltonian}
	A continuous deterministic and reversible process for a homogeneous classical material admits a Hamiltonian $H: T^*\mathcal{Q} \rightarrow \mathbbm{R}$ that allows us to write the laws of evolution as
\begin{align*}
d_{t}q^i &= \partial_{p_i} H \\
d_{t}p_i &= - \partial_{q^i} H
\end{align*}
\end{prop}

@@

We claim the state variables evolve according to Hamilton's equations. $S^{a} \omega_{ab} = d_t\xi^a \omega_{ab} = \partial_{b}H$. For $b=\{1,...,n\}$ we have $d_tp_i (-1) = \partial_{q_i} H$. For $b=\{n+1,...,2n\}$ we have $d_tq^i (+1) = \partial_{p_i} H$.
\end{justification}

We recognize the familiar set of Hamilton's equations. They are the set of equations to describe the deterministic and reversible motion of the infinitesimal parts of an infinitesimally reducible homogeneous material. In other words: the forces that conserve energy (i.e. the value of a suitable Hamiltonian) are exactly the ones that provide deterministic and reversible motion. The challenge in their derivation mainly lies in the necessary use of different branches of mathematics (e.g. topology, measure theory, differential geometry, symplectic geometry). The conceptual meaning, on the other hand, is hopefully straight forward.

It's important to realize that, during the derivation, multiple mathematical features (e.g. invariant densities, cotangent bundle for phase space, symplectomorphism) were justified by the same physical assumption. The math itself gives us no indication that the different features stem from the same source; therefore, the math itself does not give us a conceptually unified picture. This is one of the reasons we believe that starting from the physical description is objectively better if we are to come to a better understanding of our physical theories.

Note that we could have taken different approaches. For example, we could have appealed to information theory and required our invariant distributions to preserve Shannon's information entropy, as no information should be gained or lost during a deterministic and reversible process. Or we could have appealed to statistical mechanics and required that the determinant of the covariance matrix be conserved, as a deterministic and reversible process should be defined at the same level of precision. With suitable treatment of independent d.o.f. both these approaches recover Hamiltonian mechanics as well. While the full treatment is outside the scope of this work, we want to underline that the definitions used in this work go a long way to building bridges at the core of different mathematical and scientific branches.

@@

The discussion so far has been limited to the case where both the set of states and the evolution map are defined at equal time and never change throughout the evolution. This is too restrictive as there are very reasonable situations in which this does not hold.

The first obvious case is that the map may be time dependent. Note that this does not affect our definition of determinism and reversibility: we still can tell final state from initial state and vice-versa. We just need to specify the time as well. At this point we have no way to do that. For example, the Hamiltonian $H: T^*\mathcal{Q} \rightarrow \mathbbm{R}$ is just a function of the state.

@@

The third case is when we perform a state variable transformation $q'=q'(q,t)$ that is time dependent (e.g. $q'=q+vt$). Our composite state distribution $\rho(q)$ becomes $\rho(q',t)$: it is no longer defined at equal time. This means that both our measure $\mu$ and our metric $d\mu$ need to be modified.

Therefore we cannot simply stick time in the Hamiltonian and expect everything to work out. But we can't ignore the problem either as the three cases outlined are pretty common situations to study. So we need to go back to our definitions and amend them properly.

@@

The first thing to do is to extend our $n$-dimensional $\mathcal{Q}$, the manifold that defines our units, to include the time variable $t$. We call $\mathcal{M}$ such extended space which is a manifold of dimension $n+1$: state variables and the time variable are independent (i.e. any combination of state and time is valid). A time variable is not simply another state variable: it doesn't identify extra states and it's the variable on which we have defined deterministic and reversible evolution. When changing variables, we still need to make sure that deterministic and reversible motion can be defined on the new time variable. Transformations like $t'=q, \; q'=t$ (exchanging time with another state variable) or $t'=\sqrt{t^2 + q^2}, \; q'=\tan^{-1}(t/q)$ (polar coordinates between time and state variable) clearly do not guarantee deterministic and reversible motion over the new time variable. The change of time variable must preserve the time ordering therefore the transformation has to be strictly monotonic. In this sense, the time variable is special.

As before, we are interested in invariant densities which are defined on co-vectors. TODO: elaborate on covector component. The time variable $t$ will have a conjugate quantity $E\equiv\hbar\bar{\omega}$.\footnote{$\bar{\omega}$ is the classical analogue of the wave frequency. We use $\bar{\omega}$ to distinguish from the phase-space metric $\omega$.} We call the combination of $(t, E)$ the temporal degree of freedom. As time is special, so is the temporal degree of freedom, which is treated differently when defining distribution $\rho_\mathcal{c}$ and therefore by the measure $\mu$ and the metric $\omega$.

The first thing to note is that $E$ cannot be an independent variable. It cannot add any states, as those are defined by $T^*\mathcal{Q}$ only, and cannot add any time instants, as those are defined by $t$ alone. Therefore there must exist a constraint such that, locally, $E=E(t,q^i,p_i)$. The set of all states at all times is a sub-manifold of $T^*\mathcal{M}$ of dimension $2n+1$.

@@

We can understand the geometry of $T^*\mathcal{M}$ in the following way. As we have seen, the temporal degree of freedom $(t, E)$ and the other d.o.f. $(q^i, p_i)$ are not independent. So they are not orthogonal in the extended phase space. States are defined on the plane where $( q, p )$ (maximally) change: this is not the plane of constant $( t, E )$ (they are not orthogonal) where $dq \wedge dp$ is defined, but the plane perpendicular to constant $( q, p )$ where $dt \wedge dE$ is defined. It is on that plane we can properly count states and define our metric.

@@

The metric $\omega$ is still the canonical two form expressed with the mathematically non-canonical, but physically meaningful, variable $E$. This allows us to better understand its physical meaning. The metric allows us to count the possibilities within a degree of freedom, adjusting the metric on $T^*\mathcal{Q}$ to avoid the counting problems introduced by time variable changes.

The metric actually fixes two other problems. The first is: how do we distinguish temporal from standard d.o.f.? Given a two-dimensional surface $U \subset T^*\mathcal{M}$, how can we tell if it should be charted by a $(t,E)$ or whether we can define a distribution on it? Consider $\int_U \omega$. The result will be positive if the integration is over a standard degree of freedom, and it will be negative if the integration is over the temporal degree of freedom. If the contribution in every subregion of $U$ is positive, then the whole $U$ is always oriented along a standard degree of freedom.

The second problem is: how do we prevent time variable changes that disturb determinism and reversibility? The invariance of $\omega$ already guarantees that. As we saw before that fact that the vector product is conserved within a degree of freedom is equivalent to saying that changing variables conserves the possibilities within a d.o.f. In fact, the flow of an infinitesimal transformation that preserves the vector product on a surface is a divergence free field. The conservation of the Minkowsky product across the temporal and standard d.o.f. means that changing variables does not introduce rotation between time and state variables. In fact, the flow of an infinitesimal transformation that preserves the Minkowsky product on a surface is a curl free field.

We can capture the above discussion by stating that the extended state space is a hypersurface of the symplectic manifold $(T^*\mathcal{M}, \omega)$, where $\omega \equiv \sum dq^i \wedge dp_i - dt \wedge dE$. This is the simplest manifold that allows us to define time-and-state-variable-invariant densities. The metric allows us to measure the cardinality of possibilities.

\begin{prop}\label{classical_extended_phase_space}
	The extended state space, the set of states at all possible times, for the particles of a homogeneous classical material is a hyper-surface of the symplectic manifold formed by a cotangent bundle $T^*\mathcal{M}$ equipped with the canonical two-form $\omega = \sum dq^i \wedge dp_i - dt \wedge dE$.
\end{prop}
